---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
if (!file.exists ("activity.csv"))
  unip ("activity.zip")
act <- read.csv ("activity.csv")
```

## What is mean total number of steps taken per day?

```{r}
dailyStepTotal <- aggregate (. ~ date, data = act, sum)
hist (dailyStepTotal$steps, xlab = "Total Steps per Day")
abline (v = mean (dailyStepTotal$steps), col = "red")
legend ("topright", lty = 1, bty = "n", col = "red", legend = "Mean Steps per Day")

meanStepsPerDay <- sprintf ("%.0f", mean (dailyStepTotal$steps))
medianStepsPerDay <- sprintf ("%.0f",median (dailyStepTotal$steps))
```

Mean steps per day: `r meanStepsPerDay`  
Median steps per day: `r medianStepsPerDay`


## What is the average daily activity pattern?

```{r}
stepsEachInterval <- aggregate (steps ~ interval, data = act, mean)
with (stepsEachInterval, plot (interval, steps, type = "l", 
      xlab = "5 Minute Intervals Throughout Each Day", ylab = "Steps"))

rowNum <- which (stepsEachInterval$steps == max (stepsEachInterval$steps))
intervalWithMaxSteps <- stepsEachInterval$interval[rowNum]

abline (v = intervalWithMaxSteps, col = "blue")
legend ("topright", lty = 1, bty = "n", col = "blue", legend = "Interval With Max Steps")

```

The interval number with the maximum average number of steps is: `r intervalWithMaxSteps`.

## Imputing missing values

```{r}
numMissingValues <- sum (is.na(act$steps))
```

Total number of missing values: `r numMissingValues`.  

```{r}
# Strategy for imputing missing values: use average for that interval over all days in dataset
avgLookupVector <- rep (0L, max (stepsEachInterval$interval))
for (i in 1:length (stepsEachInterval$steps)) {
  interval <- stepsEachInterval$interval[i]
  avgLookupVector[interval+1] <- stepsEachInterval$steps[i]
}
imputedAct <- act
for (i in 1:length (imputedAct$steps)) {
  if (is.na (imputedAct$steps[i])) {
      imputedAct$steps[i] <- avgLookupVector[imputedAct$interval[i] + 1]
  }
}

dailyStepTotal <- aggregate (. ~ date, data = imputedAct, sum)
hist (dailyStepTotal$steps, xlab = "Total Steps per Day")
abline (v = mean (dailyStepTotal$steps), col = "red")
legend ("topright", lty = 1, bty = "n", col = "red", legend = "Mean Steps per Day")

meanStepsPerDay <- sprintf ("%.0f", mean (dailyStepTotal$steps))
medianStepsPerDay <- sprintf ("%.0f",median (dailyStepTotal$steps))
```

Mean steps per day, using imputed missing values: `r meanStepsPerDay`  
Median steps per day, using imputed missing values: `r medianStepsPerDay`




## Are there differences in activity patterns between weekdays and weekends?
